A Partitioning Algorithm for Maximum Common (Connected) Subgraphs\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/K503058/1 and EP/M508056/1]}

### 2017-02-10 paper/paper.tex
More algorithm-writing

@@

@@

\begin{algorithm}
\DontPrintSemicolon
\nl $\FuncSty{expand}(\mathit{future},matching)$ \;
\nl \Begin{
\nl \lIf {$future = \emptyset$ \bf{and} $|matching|  > |incumbent|$}{$save(matching)$}
\nl \lIf {$future = \emptyset$}{return}
\nl $bestFuture \gets 0$ \;
\nl \lFor {$\langle P,T \rangle \in future$}{$bestFuture \gets bestFuture + min(|P|,|T|)$}
\nl \lIf {$|matching|  + bestFuture \leq |incumbent|$}{return}
\bigskip
\nl $\langle P,T \rangle \gets select(future)$ \;
\nl $v \gets select(P)$ \;
\nl \For {$w \in T$}{
\nl    $matching \gets matching + (v,w)$\;
\nl    \bf{Let} $future^{'} \gets \emptyset$ \;
\nl    \For {$\langle P^{'},T'^{'}\rangle \in future$}{
\nl        \bf{Let} $P^{''} \gets P^{'} \cap N(v,G) \setminus \{v\}$ \;
\nl        \bf{Let} $T^{''} \gets T^{'} \cap N(w,H) \setminus \{w\}$ \;
\nl        \lIf {$P^{''} \neq \emptyset$ \bf{and} $T^{''} \neq \emptyset$}{$future^{'} \gets future^{'} + \langle P^{''} , T^{''} \rangle$}
\nl        \bf{Let} $P^{''} \gets P^{'} \cap N(v,\overline{G}) \setminus \{v\}$ \;
\nl        \bf{Let} $T^{''} \gets T^{'} \cap N(w,\overline{H}) \setminus \{w\}$ \;
\nl        \lIf {$P^{''} \neq \emptyset$ \bf{and} $T^{''} \neq \emptyset$}{$future^{'} \gets future^{'} + \langle P^{''} , T^{''} \rangle$}
       }
\nl   $\FuncSty{expand}(future^{'},matching)$ \;
\nl   $matching \gets matching \setminus \{(v,w)\}$ \;
  }
\nl $P \gets P \setminus \{v\}$\;
\nl \lIf {$P = \emptyset$} {$future \gets future \setminus \{\langle P,T \rangle \}$}
\nl $\FuncSty{expand}(future,matching)$ \;
}
\;
\nl $\FuncSty{McSplit}(G,H)$ \;
\nl \Begin{
\nl $incumbent \gets \emptyset$ \;
\nl $\FuncSty{expand}(\{\langle V(G),V(H) \rangle \},\emptyset)$ \;
\nl return $|incubent|$ \;
}
\caption{\McSplit\ algorithm for Maximum Common Subgraph}
\label{jtAlg}
\end{algorithm}

The full algorithm is shown as Algorithm~\ref{jtAlg}, with input graphs $G$ and $H$.
The main recursive procedure, future is a list of pairs, where the first element of a pair
is a set of pattern vertices P and the second is a set of target vertices T. It
is assumed that each pattern vertex in P might be matched with any (all) of the
target vertices in T, or be unmatched. Therefore we start our search via the
function LilybankSplit (line 26). This returns an integer, the size of the
largest matching.  In line 29 a call is made to expand, where we have as
arguments a list of pairs (called future, line 1) and a set of pairs of the
form (v,w) where pattern vertex v is matched to target vertex w.

We now focus on expand (line (1)). This is initially called (line 29) with the
future as a list containing the single pair $\langle V(G),V(H) \rangle$. Lines
4 to 7 define the conditions where we return, i.e. when the incumbent cannot be
improved. In line 3, a new incumbent might be saved. Lines 8 to 24 perform the
search.

A pair is selected from the future (line 8) using some heuristic. Note that
this pair is not removed from the future, and that P and T can be considered as
pointers to sets. A pattern vertex is then selected, possibly using a heuristic
(line 9). We now iterate over all possible target vertices in T (lines 10 to
21). A target vertex w is selected and a matching produced (line 11). A new
future is created (using {\bf Let} to signify a new variable, line 12). Every
pair in the future can now be split (lines 13 to 19) into pattern vertices
adjacent to the selected pattern vertex v and target vertices adjacent to the
selected target vertex w (lines 14 to 15) and this is added to the new future
if both sets are not null (line 16) i.e. there is a potential for more
matchings. This is then repeated symmetrically for non-adjacency (lines 17 to
19) using the compliments of the graphs.  A recursive call is then made using
the new future and the increased matching (line 20). On return from that call
we remove the matching (v,w) (line 21). Having explored all matchings with v in
its target set we now consider what happens if the selected pattern vertex v is
not matched (line 22 to 24), where that set of matchings does not contain a
pair (v,*).

>>>>

\newcommand{\AlgVar}[1]{\mathit{#1}}

@@

\newcommand{\lineref}[1]{line~\ref{#1}}
\newcommand{\twolinesref}[2]{lines~\ref{#1} and~\ref{#2}}
\newcommand{\Lineref}[1]{Line~\ref{#1}}
\newcommand{\Twolinesref}[2]{Lines~\ref{#1} and~\ref{#2}}

@@

\begin{algorithm}
\DontPrintSemicolon
\nl $\FuncSty{expand}(\AlgVar{future},M)$ \;
\nl \Begin{
%\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
\nl \lIf {$|M| > |\AlgVar{incumbent}|$}
    {$save(matching)$} \label{StoreIncumbent}
%\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
\nl $bestFuture \gets 0$ \label{SetBestFutureToZero} \;
\nl \lFor {$\langle P,T \rangle \in \AlgVar{future}$}
    {$\AlgVar{bestFuture} \gets \AlgVar{bestFuture} + \min(|P|,|T|)$}
\nl \lIf {$|M|  + \AlgVar{bestFuture} \leq |\AlgVar{incumbent}|$}{return} \label{PruneSearch}
\bigskip
\nl $\langle P,T \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
\nl $v \gets \FuncSty{SelectVertex}(P)$ \label{SelectVertex} \;
\nl \For {$w \in T$ \label{WLoop}} {
\nl    $M \gets M \cup (v,w)$ \label{GrowM} \;
\nl    \bf{Let} $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
\nl    \For {$\langle P',T'\rangle \in future$ \label{InnerLoop}}{
\nl        \bf{Let} $P'' \gets P' \cap N(v,G) \setminus \{v\}$ \label{NewPWithEdge} \;
\nl        \bf{Let} $T'' \gets T' \cap N(w,H) \setminus \{w\}$ \;
\nl        \lIf {$P'' \neq \emptyset$ \bf{and} $T'' \neq \emptyset$}
        {$\AlgVar{future'} \gets \AlgVar{future'} + \langle P'' , T'' \rangle$ \label{AddToFutureWithEdge}}
\nl        \bf{Let} $P'' \gets P' \cap N(v,\overline{G}) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
\nl        \bf{Let} $T'' \gets T' \cap N(w,\overline{H}) \setminus \{w\}$ \;
\nl        \lIf {$P'' \neq \emptyset$ \bf{and} $T'' \neq \emptyset$}
        {$\AlgVar{future'} \gets \AlgVar{future'} + \langle P'' , T'' \rangle$} \label{InnerLoopEnd}
       }
\nl   $\FuncSty{expand}(\AlgVar{future'},M)$ \label{ExpandWithV} \;
\nl   $M \gets M \setminus \{(v,w)\}$ \label{ShrinkM} \;
  }
\nl $P' \gets P \setminus \{v\}$ \label{RemoveV} \;
\nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle P,T \rangle\} \cup \{\langle P', T \rangle\}$\;
\nl \lIf {$P' = \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle P',T \rangle \}$}
\nl $\FuncSty{expand}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
}
\;
\nl $\FuncSty{McSplit}(G,H)$ \label{McSplitFun} \;
\nl \Begin{
\nl $\AlgVar{incumbent} \gets \emptyset$ \;
\nl $\FuncSty{expand}(\{\langle V(G),V(H) \rangle \},\emptyset)$ \label{FirstExpandCall} \;
\nl return $|\AlgVar{incubent}|$ \;
}
\caption{\McSplit\ algorithm for Maximum Common Subgraph}
\label{jtAlg}
\end{algorithm}

\paragraph{Full algorithm} The full algorithm is shown as
Algorithm~\ref{jtAlg}.  The main recursive procedure, $\FuncSty{expand}$, has
two parameters.  The parameter $\AlgVar{future}$ is a list of label classes,
each represented as a $\langle P, T \rangle$ pair as described above.  The
parameter $M$ is the current mapping of vertices.  On each call to
$\FuncSty{expand}$, the invariant holds that a $(v,w)$ pair may be added to $M$
if and only if $v$ and $w$ belong to the same label class in $M$.

\Lineref{StoreIncumbent} stores the current mapping $M$ if it is large enough
to unseat the incumbent.  \Twolinesref{SetBestFutureToZero}{PruneSearch} prune
the search when a calculated upper bound ($\AlgVar{bestFuture}$) is not larger
than the incumbent.

The remainder of $\FuncSty{expand}$ performs the search.  A label-class
$\langle P, T \rangle$ is selected from $\AlgVar{future}$
(\lineref{SelectClass}) using some heuristic.  From this class, a pattern
vertex $v$ is selected using a heuristic (\lineref{SelectVertex}). We now
iterate over all possible vertices in $T$ (\twolinesref{WLoop}{ShrinkM}). A
vertex $w \in \V(H)$ is selected mapping $M$ is extended (\lineref{GrowM}). A
new set of label-classes, $\AlgVar{future'}$, is created (\lineref{NewFuture}).
Every label-class in $\AlgVar{future}$ can now be split
(\twolinesref{InnerLoop}{InnerLoopEnd}) into two new classes. The first of
these classes (\twolinesref{NewPWithEdge}{AddToFutureWithEdge}) contains
pattern vertices adjacent to $v$ and target vertices adjacent to $w$.  This is
added to $\AlgVar{future'}$ if both sets contain at least one vertex.  This is
then repeated symmetrically for non-adjacency
(\twolinesref{NewPWithoutEdge}{InnerLoopEnd}) using the compliments of the
graphs.  A recursive call is then made (\lineref{ExpandWithV}), on return from
which we remove the mapping $(v,w)$.  Having explored all possible mappings of
$v$ its target set we now consider what happens if $v$ is not matched
(\twolinesref{RemoveV}{ExpandWithoutV}). Note that removing $v$ from its label
class is equivalent in effect to assigning it the label $\bot$, which is never
taken by any vertex in $\V(H)$.

We start our search at the function $\FuncSty{McSplit}$ (\lineref{McSplitFun}),
with graphs $G$ and $H$ as inputs.  This function returns the size of the
largest mapping.  In \lineref{FirstExpandCall} a call is made to
$\FuncSty{expand}$; at this point we have a single label-class containing all
vertices, and the mapping $M$ is empty.
